---
title: "Predicting daily Citibike trips (open-ended)"
author: "Basira Shirzad"
date: "'`r Sys.time()`"
output: 
  html_document:
    toc: yes
    toc_depth: '3'
    df_print: paged
  pdf_document: 
    toc: yes
    toc_depth: 3
---

# Setup

Load some packages to help us model our data
```{r setup}
library(tidyverse)
library(scales)
library(modelr)
library(lubridate)

# set plot theme
theme_set(theme_bw())
options(repr.plot.width=4, repr.plot.height=3)
knitr::opts_chunk$set(echo = TRUE)
```

# Load and preview data

Read data from the `trips_per_day.tsv` file

```{r load-trip-data}
trips_per_day <- read_tsv('trips_per_day.tsv')
head(trips_per_day)
```
Read data from the `US_Bank_holidays` file

```{r load-holiday-data}
holidays <- read_csv('US_Bank_holidays', col_names = c('day_num', 'ymd', 'holiday_name'))
head(holidays)
```
Create a model_data frame that includes information from both files 

```{r load-new-data}
model_data <- trips_per_day %>% 
  mutate(day = as.factor(wday(ymd, label = F)), is_weekday = between(wday(ymd),2,6), is_rain = prcp > 0) %>%
  left_join(holidays) %>%
  mutate(is_holiday = !is.na(holiday_name)) %>%
  select(ymd, num_trips, prcp, tmax, tmin, day, is_weekday, is_rain, is_holiday)
head(model_data)
```
# Plot Graph

Plot for the number of trips taken as a function of the minimum temperature on each day.

```{r weather-tmin-plot}
ggplot(model_data, aes(x = tmin, y = num_trips)) +
  geom_point(aes(colour = day)) +
  xlab('Minimum temperature') +
  ylab('Daily trips') +
  scale_y_continuous()

ggplot(model_data, aes(x = tmin, y = num_trips)) +
  geom_point(aes(colour = is_weekday)) +
  xlab('Minimum temperature') +
  ylab('Daily trips') +
  scale_y_continuous()

ggplot(model_data, aes(x = tmin, y = num_trips)) +
  geom_point(aes(colour = is_rain)) +
  xlab('Minimum temperature') +
  ylab('Daily trips') +
  scale_y_continuous()

ggplot(model_data, aes(x = tmin, y = num_trips)) +
  geom_point(aes(colour = is_holiday)) +
  xlab('Minimum temperature') +
  ylab('Daily trips') +
  scale_y_continuous()
```

Plot for the number of trips taken as a function of the maximum temperature on each day.

```{r weather-tmax-plot}
ggplot(model_data, aes(x = tmax, y = num_trips)) +
  geom_point(aes(colour = day)) +
  xlab('Maximum temperature') +
  ylab('Daily trips') +
  scale_y_continuous()

ggplot(model_data, aes(x = tmax, y = num_trips)) +
  geom_point(aes(colour = is_weekday)) +
  xlab('Maximum temperature') +
  ylab('Daily trips') +
  scale_y_continuous()

ggplot(model_data, aes(x = tmax, y = num_trips)) +
  geom_point(aes(colour = is_rain)) +
  xlab('Maximum temperature') +
  ylab('Daily trips') +
  scale_y_continuous()

ggplot(model_data, aes(x = tmax, y = num_trips)) +
  geom_point(aes(colour = is_holiday)) +
  xlab('Maximum temperature') +
  ylab('Daily trips') +
  scale_y_continuous()
```

# Cross-validation

Fitting different polynomials to the data and using cross-validation (shuffled data and used 80% of the data for training and 10% of the data for validation) to find the polynomial degree that generalizes best to held out data.

```{r cross-validation}
# Random number generator that produces the same sequence of numbers
set.seed(42)

num_days <- nrow(model_data)
frac_train <- 0.8 # train on 80% of the data
num_train <- floor(num_days * frac_train)

# randomly sample rows for the training set 
ndx <- sample(1:num_days, num_train, replace=F)

# used to fit the model
trips_per_day_train <- model_data[ndx, ] 

# used to evaluate the fit
trips_per_day_remaining <- model_data[-ndx, ]
num_remaining = floor(nrow(trips_per_day_remaining)/2)

ndx2 <- sample(1:nrow(trips_per_day_remaining), num_remaining, replace = F)
trips_per_day_validate <- trips_per_day_remaining[ndx2, ]
trips_per_day_test <- trips_per_day_remaining[-ndx2, ] 
```

# Evaluation

Evaluating models from degree 1 to degree 8, fitting on the training data and evaluate on the validation data.

```{r evaluate-models}
# fit a model for each polynomial degree
K <- 1:8
train_err <- c()
validate_err <- c()
for (k in K) {
  
    # fit on the training data
    model <- lm(num_trips ~ poly(tmin, k, raw = T) + poly(tmax, k, raw = T) + day + is_weekday + is_rain + is_holiday, data=trips_per_day_train)
    
    # evaluate on the training data
    train_err[k] <- sqrt(mean((predict(model, trips_per_day_train) - trips_per_day_train$num_trips)^2))

    # evaluate on the validate data
    validate_err[k] <- sqrt(mean((predict(model, trips_per_day_validate) - trips_per_day_validate$num_trips)^2))
}

summary(model)
```

Plot for the training and validation error as a function of the polynomial degree.

```{r plot-errors}
plot_data <- data.frame(K, train_err, validate_err) %>%
  gather("split", "error", -K)

ggplot(plot_data, aes(x=K, y=error, color=split)) +
  geom_line() +
  scale_x_continuous(breaks=K) +
  xlab('Polynomial Degree') +
  ylab('RMSE')
```

Fitting and plotting the model on all of the data 

```{r test-data}
# chose final k
k = 4
model <- lm(num_trips ~ poly(tmin, k, raw = T) + poly(tmax, k, raw = T) + day + is_weekday + is_rain + is_holiday, data=trips_per_day_train)

trips_per_day_train <- trips_per_day_train %>%
  add_predictions(model) %>%
  mutate(split = "train")
trips_per_day_validate <- trips_per_day_validate %>%
  add_predictions(model) %>%
  mutate(split = "validate")
plot_data <- bind_rows(trips_per_day_train, trips_per_day_validate)

ggplot(plot_data, aes(x = tmax, y = num_trips)) +
  geom_point(aes(color = split)) +
  geom_line(aes(y = pred)) +
  geom_smooth(se = FALSE) +
  xlab('Maximum temperature') +
  ylab('Daily trips') +
  scale_y_continuous()
```
```{r results}
rsquare(model, plot_data)
rmse(model, plot_data)
```

# Re-Evaluating 

Re-evaluating models from degree 1 to degree 8, fitting on the training data and evaluate on the validation data.

```{r re-evaluate-models}
# fit a model for each polynomial degree
K <- 1:8
train_err <- c()
validate_err <- c()
for (k in K) {
  
    # fit on the training data
    model <- lm(num_trips ~ poly(tmax, k, raw = T) + day + is_rain + is_holiday, data=trips_per_day_train)
    
    # evaluate on the training data
    train_err[k] <- sqrt(mean((predict(model, trips_per_day_train) - trips_per_day_train$num_trips)^2))

    # evaluate on the validate data
    validate_err[k] <- sqrt(mean((predict(model, trips_per_day_validate) - trips_per_day_validate$num_trips)^2))
}

summary(model)
```

Re-Plot for the training and validation error as a function of the polynomial degree.

```{r re-plot-errors}
plot_data <- data.frame(K, train_err, validate_err) %>%
  gather("split", "error", -K)

ggplot(plot_data, aes(x=K, y=error, color=split)) +
  geom_line() +
  scale_x_continuous(breaks=K) +
  xlab('Polynomial Degree') +
  ylab('RMSE')
```

Re-fitting and plotting the model on all of the data 

```{r re-test-data}
# chose final k
k = 5
model <- lm(num_trips ~ poly(tmax, k, raw = T) + day + is_rain + is_holiday, data=trips_per_day_train)

trips_per_day_train <- trips_per_day_train %>%
  add_predictions(model) %>%
  mutate(split = "train")
trips_per_day_validate <- trips_per_day_validate %>%
  add_predictions(model) %>%
  mutate(split = "validate")
plot_data <- bind_rows(trips_per_day_train, trips_per_day_validate)

ggplot(plot_data, aes(x = tmax, y = num_trips)) +
  geom_point(aes(color = split)) +
  geom_line(aes(y = pred)) +
  geom_smooth(se = FALSE) +
  xlab('Maximum temperature') +
  ylab('Daily trips') +
  scale_y_continuous()
```
```{r final-results}
rsquare(model, plot_data)
rmse(model, plot_data)
```

# Final Best Fit Model

```{r graph-1}
ggplot(plot_data, aes(x = ymd, y = num_trips)) +
  geom_point(aes(color = split)) +
  geom_line(aes(y = pred)) +
  geom_smooth(se = FALSE) +
  xlab('Date') +
  ylab('Predicted Trips (red) & Actual Trips (black)') +
  scale_y_continuous()

ggplot(plot_data, aes(x = pred, y = num_trips)) +
  geom_point(aes(color = split)) +
  geom_abline() +
  xlab('Predicted Trips Count') +
  ylab('Actual Trip Count') +
  scale_y_continuous()
```

# Testing on Test Data

```{r test-dataset}

# chose final k

trips_per_day_test <- trips_per_day_test %>%
  add_predictions(model) %>%

ggplot(trips_per_day_test, aes(x = tmax, y = num_trips)) +
  geom_point(aes(color = split)) +
  geom_line(aes(y = pred)) +
  geom_smooth(se = FALSE) +
  xlab('Maximum temperature') +
  ylab('Daily trips') +
  scale_y_continuous()

# Results
rsquare(model, plot_data)
rmse(model, plot_data)

ggplot(trips_per_day_test, aes(x = ymd, y = num_trips)) +
  geom_point(aes(color = split)) +
  geom_line(aes(y = pred)) +
  geom_smooth(se = FALSE) +
  xlab('Date') +
  ylab('Predicted Trips (red) & Actual Trips (black)') +
  scale_y_continuous()

ggplot(trips_per_day_test, aes(x = pred, y = num_trips)) +
  geom_point(aes(color = split)) +
  geom_abline() +
  xlab('Predicted Trips Count') +
  ylab('Actual Trip Count') +
  scale_y_continuous()
```

